Neural and Symbolic Spatiotemporal Optimization of Cosmological Particel-Mesh Simulation

### 2023-02-28 docs/papers/sto/sto.tex
FIXUP draft: shrink Ob/Om and Ok/(1-Ok) ranges
FIXUP generate IC for g4 on GPU
update TODO

\section*{TODO}
\begin{itemize}
\item Compare one pair of Gadget4 and \pmwd\ (with say 100 to 1000 time
    steps) results and make sure that they agree perfectly on the large
    scales. For this purpose the box size should be large enough (in
    linear regime), say 1 Gpc/$h$.
\item Be careful with the non-deterministic ICs generated on GPUs.
  \begin{enumerate}
  \item Saving them on disks and loading them during training solves
    this but may introduce IO or memory problem;
  \item Generating them on CPUs for training data and also training
    might suffice while requireing such data loader implementation;
  \item Generating them on GPUs for training data and training will
    introduce additional source of initial deviations to be nonlinearly
    amplified. There are flags that makes certain operations
    deterministic but slow on GPUs, but those are enviroment variables
    that function globally and cannot be applied locally only on the IC
    generation during training.
  \item Generating ICs and then train on GPUs with double precision and
    then compute the loss at single precision (Gadget4 should be run at
    double precision) should work, and likely be the simplest solution.
    But for this to work the final training loss should be much larger
    than the level of reproducibility.
  \end{enumerate}
\item Non-integer multiple mesh shapes: for uniform particle grid, test
  (with double precision on CPUs) whether they start moving by artifact
  forces on cell scales, in 3 cases: no deconvolution, with CIC
  deconvolution, with SO.
\item For TO, probably necessary to first incorporate Oliver Hahn's
  recent improvement on FastPM integrator arXiv:2301:09655.
\end{itemize}

@@

\begin{table*}
\centering
\caption{Ranges of GADGET-4 and \pmwd\ configuration and cosmological
parameters.
Note that the grid ratio need next fast len to determine the mesh shape
for fast FFT.
Given the box size, the mesh shape determines the cell size.
The softening parameter gives the ratio of the comoving softening length
to the mean particle spacing.
The curvature $\Omega_k$ is related to the separate universe simulation.
We sample parameters applicable to GADGET-4 during data generation, and
those applicable to \pmwd\ during training.
\textsuperscript\dag The box size is determined jointly by the \pmwd\
mesh shape and mesh cell size below, which are assumed to be sampled
independently.
In practice, we sample one conditioned on the other and the box size.
* Number of time steps from $a=1/16$ to $a=1$.
}
\label{tab:param}
\begin{tabular}{lcccr}
  \toprule
  applicability & parameter & distribution & range & sampling \\
  \midrule
  & $128^3$ white noise & $\mathcal{N}(\vzero, \vI)$ & 512 realizations & MC \\
  \cmidrule(lr){2-5}
  & 121 snapshots, $a$ & uniform & [1/16, 1] & deterministic \\
  \cmidrule(lr){2-5}
  & box size in Mpc\textsuperscript\dag & log-trapezoidal & [25.6, 2560) \\
  \cmidrule(lr){2-4}
  & snapshot offset, $\Delta\!a$ & uniform & [0, 1/128) \\
  \cmidrule(lr){2-4}
  GADGET-4 & $\As \times 10^9$ & log-uniform & [1, 4) \\
  \cmidrule(lr){2-4}
  \& \pmwd\ & $\ns$ & log-uniform & [0.75, 1.25) & RQMC \\
  \cmidrule(lr){2-4}
  & $\Omegam$ & log-uniform & [1/5, 1/2) & see \autoref{fig:sobol} \\
  \cmidrule(lr){2-4}
  & $\Omegab / \Omegam$ & log-uniform & [1/16, 1/2) \\
  \cmidrule(lr){2-4}
  & $\Omega_k / (1 - \Omega_k)$ & uniform & [-0.5, 0.5) \\
  \cmidrule(lr){2-4}
  & $h$ & log-uniform & [0.5, 1) \\
  \cmidrule(lr){1-4}
  GADGET-4 & softening ratio & log-uniform & [1/50, 1/20) \\
  \cmidrule(lr){1-5}
  & mesh shape & log-uniform & [128, 512] \\
  \cmidrule(lr){2-4}
  \pmwd\ & mesh cell size in Mpc & log-uniform & [0.2, 5] & MC \\
  \cmidrule(lr){2-4}
  & number of time steps* & log-uniform & [10, 1000] \\
  \bottomrule
\end{tabular}
\end{table*}

>>>>

\section*{TODO}
\begin{itemize}
\item Compare one pair of Gadget4 and \pmwd\ (with say 100 to 1000 time
    steps) results and make sure that they agree perfectly on the large
    scales. For this purpose the box size should be large enough (in
    linear regime), say 1 Gpc/$h$.
    \textbf{Update: for Sobol=6, boxsize=983Mpc, Gadget4 and \pmwd\ are visually
    consistent at the field level. To further check power spectrum.}
\item Be careful with the non-deterministic ICs generated on GPUs.
      \textbf{Update: we have shown that ICs generated on GPUs are deterministic
      (but different from those on CPUs), generating ICs for both training data
      and training on GPUs might be the simplest solution.}
  \begin{enumerate}
  \item Saving them on disks and loading them during training solves
    this but may introduce IO or memory problem;
  \item Generating them on CPUs for training data and also training
    might suffice while requireing such data loader implementation;
  \item Generating them on GPUs for training data and training will
    introduce additional source of initial deviations to be nonlinearly
    amplified. There are flags that makes certain operations
    deterministic but slow on GPUs, but those are enviroment variables
    that function globally and cannot be applied locally only on the IC
    generation during training.
  \item Generating ICs and then train on GPUs with double precision and
    then compute the loss at single precision (Gadget4 should be run at
    double precision) should work, and likely be the simplest solution.
    But for this to work the final training loss should be much larger
    than the level of reproducibility.
  \end{enumerate}
\item Non-integer multiple mesh shapes: for uniform particle grid, test
  (with double precision on CPUs) whether they start moving by artifact
  forces on cell scales, in 3 cases: no deconvolution, with CIC
  deconvolution, with SO.
\item For TO, probably necessary to first incorporate Oliver Hahn's
  recent improvement on FastPM integrator arXiv:2301:09655.
\end{itemize}

@@

\begin{table*}
\centering
\caption{Ranges of GADGET-4 and \pmwd\ configuration and cosmological
parameters.
Note that the grid ratio need next fast len to determine the mesh shape
for fast FFT.
Given the box size, the mesh shape determines the cell size.
The softening parameter gives the ratio of the comoving softening length
to the mean particle spacing.
The curvature $\Omega_k$ is related to the separate universe simulation.
We sample parameters applicable to GADGET-4 during data generation, and
those applicable to \pmwd\ during training.
\textsuperscript\dag The box size is determined jointly by the \pmwd\
mesh shape and mesh cell size below, which are assumed to be sampled
independently.
In practice, we sample one conditioned on the other and the box size.
* Number of time steps from $a=1/16$ to $a=1$.
}
\label{tab:param}
\begin{tabular}{lcccr}
  \toprule
  applicability & parameter & distribution & range & sampling \\
  \midrule
  & $128^3$ white noise & $\mathcal{N}(\vzero, \vI)$ & 512 realizations & MC \\
  \cmidrule(lr){2-5}
  & 121 snapshots, $a$ & uniform & [1/16, 1] & deterministic \\
  \cmidrule(lr){2-5}
  & box size in Mpc\textsuperscript\dag & log-trapezoidal & [25.6, 2560) \\
  \cmidrule(lr){2-4}
  & snapshot offset, $\Delta\!a$ & uniform & [0, 1/128) \\
  \cmidrule(lr){2-4}
  GADGET-4 & $\As \times 10^9$ & log-uniform & [1, 4) \\
  \cmidrule(lr){2-4}
  \& \pmwd\ & $\ns$ & log-uniform & [0.75, 1.25) & RQMC \\
  \cmidrule(lr){2-4}
  & $\Omegam$ & log-uniform & [1/5, 1/2) & see \autoref{fig:sobol} \\
  \cmidrule(lr){2-4}
  & $\Omegab / \Omegam$ & log-uniform & [1/8, 1/4) \\
  \cmidrule(lr){2-4}
  & $\Omega_k / (1 - \Omega_k)$ & uniform & [-1/3, 1/3) \\
  \cmidrule(lr){2-4}
  & $h$ & log-uniform & [0.5, 1) \\
  \cmidrule(lr){1-4}
  GADGET-4 & softening ratio & log-uniform & [1/50, 1/20) \\
  \cmidrule(lr){1-5}
  & mesh shape & log-uniform & [128, 512] \\
  \cmidrule(lr){2-4}
  \pmwd\ & mesh cell size in Mpc & log-uniform & [0.2, 5] & MC \\
  \cmidrule(lr){2-4}
  & number of time steps* & log-uniform & [10, 1000] \\
  \bottomrule
\end{tabular}
\end{table*}
